---
title: "Regression Analysis on COVID, Health, and Population Metrics"
author: "Antonio R, Aditya M, Aadam L, Christopher L"
fontsize: 12pt
geometry: "margin=2.54cm,top=1cm"
mainfont: Times New Roman
header-includes:
    - \usepackage{setspace}\singlespacing
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We can answer multiple questions using regression analysis of the covid data.
The statistical relation between total death in a country and various parameters like vaccination, age, population, etc using regression analysis of the covid dataset.
For example, We could find the correlation between total death due to Covid in a country and number of vaccinated people.

This dataset includes data on confirmed COVID cases, deaths, hospitalizations, testing, and vaccinations across
several countries since January 4, 2021. The goal of this project is to explore how a the number of new vaccinations
varies with respect to other measures over time and across different counties.
  
An individual observation corresponds to the number of new covid cases, total number of covid cases, total number of death, and
other covid, health, and population related metrics for a country in a single day since January 4, 2021. 
  
We could select multiple combinations of responses and observations to do a regression, so there doesn't need to be a specific response. However, we will focus on the total number of deaths/cases as well as total number of new deaths/cases
  
Our covariates are: days after January 4th, 2021; new tests; total tests; positive rate; tests per case; total vaccinations; people vaccinated; people fully vaccinated; new vaccinations; population; population density; median age; percentage of population aged over 65/70; GPD per capita; cardiovascular death rate; diabetes prevalence; female smokers; male smokers; hospital beds per thousand; life expectancy; human development index; stringency index.

The number of observations is 1107

```{r, echo=FALSE, out.width="50%"}
covid <- read.csv(file = 'covid.csv')
hist(covid$new_deaths, 
     breaks = 20,
     main = "Fig1: Histogram of New death due to COVID",
     axes = TRUE,
     xlab = "New Deaths",
     col = "blue")
hist(covid$total_deaths, 
     breaks = 20,
     main = "Fig2: Histogram of Total death due to COVID",
     axes = TRUE,
     xlab = "Total Deaths",
     col = "blue")
```
```{r}
show(covid)
```

```{r}
for (col in 2:ncol(covid)) {
  if(col != 11) {
    plot(covid[,col], covid[,6],
     xlab = colnames(covid)[col], ylab = "New Deaths")
  }
}
```

```{r}
date <- covid$date
approx_percentage_new_positive_tests <- ((covid$new_tests*covid$positive_rate)/covid$population)*100
approx_percentage_total_positve_tests <- ((covid$total_tests*covid$positive_rate)/covid$population)*100
tests_per_case <- covid$tests_per_case
percentage_vaccinated <- (covid$people_vaccinated/covid$population)*100
percentage_fully_vaccinated <- (covid$people_fully_vaccinated/covid$population)*100
percentage_new_vaccinated <- (covid$new_vaccinations/covid$population)*100
population_density <- covid$population_density
percentage_population_over_65 <- (covid$aged_65_older/covid$population)*100
gdp_per_capita <- covid$gdp_per_capita
percentage_cardiovascular_death <- covid$cardiovasc_death_rate/1000
diabetes_prevalence <- covid$diabetes_prevalence
hospital_beds_per_100 <- covid$hospital_beds_per_thousand/10
life_expectancy <- covid$life_expectancy
human_development_index <- covid$human_development_index
stringency_index <- covid$stringency_index
percentage_total_deaths <- (covid$total_deaths/covid$population)*100
perecentage_new_deaths <- (covid$new_deaths/covid$population)*100
temp <- cbind.data.frame(covid$location, covid$population)
agg <- aggregate(covid$new_deaths, by=list(covid$location), mean)
for(con in unique(covid$location)) {
  temp[,2][temp[,1] == con] = (agg[,2][agg[,1] == con]/temp[,2][temp[,1] == con][1])*100
}
percentage_avg_new_deaths_by_country = temp[,2]
covid_new = cbind.data.frame(date, approx_percentage_new_positive_tests, approx_percentage_total_positve_tests, tests_per_case, percentage_vaccinated, percentage_fully_vaccinated, percentage_new_vaccinated, population_density, percentage_population_over_65, gdp_per_capita, percentage_cardiovascular_death, diabetes_prevalence, hospital_beds_per_100, life_expectancy, human_development_index, stringency_index, percentage_avg_new_deaths_by_country, percentage_total_deaths, perecentage_new_deaths)
show(covid_new)
write.csv(covid_new, "C:/Users/AADAM/Google Drive/Covid525Project/new_covid.csv",row.names= FALSE)
```

```{r}
for (col in 1:ncol(covid_new)) {
  plot(covid_new[,col], covid_new$perecentage_new_deaths,
     xlab = colnames(covid_new)[col], ylab = "% New Deaths")
}
```