---
title: "Regression Analysis on COVID, Health, and Population Metrics"
author: "Antonio R, Aditya M, Aadam L, Christopher L"
fontsize: 12pt
geometry: "margin=2.54cm,top=1cm"
mainfont: Times New Roman
header-includes:
    - \usepackage{setspace}\singlespacing
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, out.width="50%"}
covid <- read.csv(file = 'covid.csv')
```

```{r}
location <- covid$location
date <- covid$date
approx_percentage_new_positive_tests <- ((covid$new_tests*covid$positive_rate)/covid$population)*100
approx_percentage_total_positve_tests <- ((covid$total_tests*covid$positive_rate)/covid$population)*100
tests_per_case <- covid$tests_per_case
percentage_vaccinated <- (covid$people_vaccinated/covid$population)*100
percentage_fully_vaccinated <- (covid$people_fully_vaccinated/covid$population)*100
percentage_new_vaccinated <- (covid$new_vaccinations/covid$population)*100
population_density <- covid$population_density
percentage_population_over_65 <- (covid$aged_65_older/covid$population)*100
gdp_per_capita <- covid$gdp_per_capita
percentage_cardiovascular_death <- covid$cardiovasc_death_rate/1000
diabetes_prevalence <- covid$diabetes_prevalence
hospital_beds_per_100 <- covid$hospital_beds_per_thousand/10
life_expectancy <- covid$life_expectancy
human_development_index <- covid$human_development_index
stringency_index <- covid$stringency_index
percentage_total_deaths <- (covid$total_deaths/covid$population)*100
perecentage_new_deaths <- (covid$new_deaths/covid$population)*100

temp <- cbind.data.frame(covid$location, covid$population)
agg <- aggregate(covid$new_deaths, by=list(covid$location), mean)
for(con in unique(covid$location)) {
  temp[,2][temp[,1] == con] = (agg[,2][agg[,1] == con]/temp[,2][temp[,1] == con][1])*100
}
percentage_avg_new_deaths_by_country = temp[,2]

covid_new = cbind.data.frame(location, date, approx_percentage_new_positive_tests, approx_percentage_total_positve_tests, tests_per_case, percentage_vaccinated, percentage_fully_vaccinated, percentage_new_vaccinated, population_density, percentage_population_over_65, gdp_per_capita, percentage_cardiovascular_death, diabetes_prevalence, hospital_beds_per_100, life_expectancy, human_development_index, stringency_index, percentage_avg_new_deaths_by_country, percentage_total_deaths, perecentage_new_deaths)
```

### Question 2
We want to find the correlation between new deaths and percentage of vaccinated population, and compare it to new deaths with respect to the percentage of fully vaccinated population.
Response: New Deaths, Covariates: Percentage of population that is vaccinated.
Response: New Deaths, Covariates: Percentage of population that is fully vaccinated.

```{r}
agg_new_deaths <- aggregate(covid_new$perecentage_new_deaths, by=list(covid_new$location), mean)$x
agg_fully_vaccinated_population <- aggregate(covid_new$percentage_fully_vaccinated, by=list(covid_new$location), mean)$x
agg_vaccinated_populations <- aggregate(covid_new$percentage_vaccinated, by=list(covid_new$location), mean)$x
```

### Observation 1
Correlation between the percentage new deaths which is the number of new deaths as a percentage of the total population and percentage of population vaccinated (one case is fully vaccinated and another in vaccinated with atleast one dose).

### Observation 2
Correlation between the percentage new deaths which is the number of new deaths as a percentage of the total population and percentage of population vaccinated (one case is fully vaccinated and another in vaccinated with atleast one dose) All aggregated by country/ location.

```{r}
boxplot(covid_new$perecentage_new_deaths,
main = "Mean of response: percentage new deaths due to Covid",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)

boxplot(covid_new$percentage_vaccinated,
main = "Mean covariate: percentage vaccinated population",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)

boxplot(covid_new$percentage_fully_vaccinated,
main = "Mean covariate: percentage fully vaccinated population",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)

boxplot(agg_new_deaths,
main = "Mean of response: percentage new deaths due to Covid, Aggregrated by country/ location",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)

boxplot(agg_vaccinated_populations,
main = "Mean covariate: percentage vaccinated population, Aggregrated by country/ location",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)

boxplot(agg_fully_vaccinated_population,
main = "Mean covariate: percentage fully vaccinated population, Aggregrated by country/ location",
xlab = "Percentage",
col = "orange",
border = "brown",
horizontal = TRUE,
notch = FALSE
)
```

### Summary of response and covariates mean,
```{r}
print("Summary for observation 1")
mean(covid_new$perecentage_new_deaths)
mean(covid_new$percentage_vaccinated)
mean(covid_new$percentage_fully_vaccinated)

print("Now observation 2, for aggregated data by country or location")
mean(agg_new_deaths)
mean(agg_vaccinated_populations)
mean(agg_fully_vaccinated_population)
```

### Summary of response and covariates standard deviation,
```{r}
print("Summary for observation 1")
sd(covid_new$perecentage_new_deaths)
sd(covid_new$percentage_vaccinated)
sd(covid_new$percentage_fully_vaccinated)

print("Now observation 2, for aggregated data by country or location")
sd(agg_new_deaths)
sd(agg_vaccinated_populations)
sd(agg_fully_vaccinated_population)
```

### Summary of response and covariates range,
```{r}
print("Summary for observation 1")
range(covid_new$perecentage_new_deaths)
range(covid_new$percentage_vaccinated)
range(covid_new$percentage_fully_vaccinated)

print("Now observation 2, for aggregated data by country or location")
range(agg_new_deaths)
range(agg_vaccinated_populations)
range(agg_fully_vaccinated_population)
```

### Scatter plots against each covariate,
```{r}
print("Summary for observation 1")
plot(covid_new$percentage_vaccinated, covid_new$perecentage_new_deaths)
plot(covid_new$percentage_fully_vaccinated, covid_new$perecentage_new_deaths)

print("Now observation 2, for aggregated data by country or location")
plot(agg_vaccinated_populations, agg_new_deaths)
plot(agg_fully_vaccinated_population, agg_new_deaths)
```

### Summary of estimated regression coefficients
```{r}
print("Summary for observation 1")
linmod <- lm(covid_new$perecentage_new_deaths ~ covid_new$percentage_vaccinated + covid_new$percentage_fully_vaccinated)

print(linmod$coefficients)
print("")

# summary(linmod)

print("Now observation 2, for aggregated data by country or location")
linmod.2 <- lm(agg_new_deaths ~ agg_vaccinated_populations + agg_fully_vaccinated_population)

print(linmod.2$coefficients)

# summary(linmod.2)
```

### Summary that includes estimated standard errors
```{r}
print("Summary for observation 1")
hist(summary(linmod)$residual)
print("R squared value")
print(summary(linmod)$r.squared)
print("MSE")
print(sigma(linmod)^2)

print("Now observation 2, for aggregated data by country or location")
hist(summary(linmod.2)$residual)
print("R squared value")
print(summary(linmod.2)$r.squared)
print("MSE")
print(sigma(linmod.2)^2)
```

### Summary that includes test results, here we found the p value and t statistic with a null hypothesis of $\beta = 0$
```{r}
print("Summary for observation 1")
n <- nrow(covid_new)
b1 <- linmod$coef[2]
s.b1 <- summary(linmod)$coef[2, 2] 
b2 <- linmod$coef[3]
s.b2 <- summary(linmod)$coef[3, 2] 
t_stat_b1 <- b1/s.b1 # Compute the test statistic of B1
t_stat_b2 <- b2/s.b2 # Compute the test statistic of B2
p <- 3
df <- n - p # Degrees of freedom
pvalue <- 2*pt(-abs(t_stat_b1), df)
print(pvalue)
pvalue <- 2*pt(-abs(t_stat_b2), df)
print(pvalue)

print("Now observation 2, for aggregated data by country or location")
n <- length(agg_new_deaths)
b1 <- linmod.2$coef[2]
b2 <- linmod.2$coef[3]
s.b1 <- summary(linmod.2)$coef[2, 2] 
s.b2 <- summary(linmod.2)$coef[3, 2] 
t_stat_b1 <- b1/s.b1 # Compute the test statistic of B1
t_stat_b2 <- b2/s.b2 # Compute the test statistic of B2
p <- 3
df <- n - p # Degrees of freedom
pvalue <- 2*pt(-abs(t_stat_b1), df)
print(pvalue)
pvalue <- 2*pt(-abs(t_stat_b2), df)
print(pvalue)
```


### TODO!!
• A short review of the key takeaways from your initial analyses. How do the analyses relate to the question(s) you
identified in IE Step 2? Is there any evidence of stronger or weaker relationships than expected?