---
title: "Regression Analysis on COVID, Health, and Population Metrics"
author: "Antonio R, Aditya M, Aadam L, Christopher L"
fontsize: 12pt
geometry: "margin=2.54cm,top=1cm"
mainfont: Times New Roman
header-includes:
    - \usepackage{setspace}\singlespacing
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, out.width="50%"}
covid <- read.csv(file = 'covid.csv')
```

```{r}

covid_new = read.csv(file='covid_new.csv')

agg_pop_death_percentage = aggregate(
    x = covid_new$percentage_total_deaths, 
    by = list(covid_new$location), 
    FUN = max
  )

agg_hdi = aggregate(
    x = covid_new$human_development_index, 
    by = list(covid_new$location), 
    FUN = mean
  )
```

## Evaluating The Model 1

The validity of the previous analysis depends on the assumptions we have made by using a regression model, so we need to verify which assumptions we violate.

```{r}

pop_death_per_country <- agg_pop_death_percentage$x
hdi <- agg_hdi$x

linmod <- lm(pop_death_per_country~hdi)
b0 <- linmod$coef[1]
b1 <- linmod$coef[2]

Yhat <- linmod$fitted.values
e <- resid(linmod)
plot(Yhat,e, xlab = expression('fitted values' ~ hat(Y)),ylab="residual e", pch = 16)
abline(h = 0, lty = 3)
```

```{r}
Y1 <- pop_death_per_country - b0
e <- resid(linmod)
plot(hdi,Y1, xlab = expression('HDI'),ylab='Y_i - b_0', pch = 16)
abline(h = 0, lty = 3)
```

```{r}
e <- resid(linmod)
par(mfrow=c(1,3))
plot(hdi, e, xlab="HDI")
hist(e, main="Histogram of residuals", xlab="residual e", col="darkmagenta", freq=FALSE)
qqnorm(e, pch = 1, frame = FALSE)
qqline(e, col = "steelblue", lwd = 2)
```

### Independence

For this regression model, the HDIs are clearly independent from each other since we have a single data point for each country in the dataset. Therefore we are not violating the independence assumption of regression models.

### Equal variance

We are also assuming equal variance in our regression model. From the residuals against the fitted values plot, we can see that they make a fan-shape, clearly, there isn't a consistent vertical spread throughout the graph. There is a moderate violation of equal variance due to this inconsistent spread. I would not suggest any transformation for the graphs, horizontally, points are evenly spread and there also aren't any outliers, which helps. For the residual against fitted values graph, using $e^{X_1}$ might help with the point concentration for higher values of $X_1$, but this really is a minor adjustment, the equal variance violation is still visible.

### Linearlity

There are no specific regions in the graph with a majority of positive or negative residuals, therefore the model doesn't exclusively underpredict or overpredict in a region, indicating that the data tends to be linear. The same can be said from the $Y_i-b_0$ against HDI.

### Normality

The QQ-Plot allows us to visualize the normality of the errors from the data. There is a moderete violation of normality in the QQ-Plot since the data is relatively evenly spread out accross the line, when more points should be centered in the middle of the line. The tails also strafe from the line.

### Review of Key Takeaways

We previously raised the question of whether the predictors defining the country's state and financial conditions actually affect the number of deaths in that country. To answer that question, we decided to use Human Development Index(HDI) to predict the total percentage of deaths in that country. We selected HDI as it takes many factors into account to give a single number describing the development of a nation. So we made a regression model that predicts the total percentage of the population of a country that died due to covid if we provide it with the HDI of the country. We found out that HDI is a decent predictor for our goal. Meaning there is a relation between countries' HDI and the percentage of people who died in that country due to covid. Detailed analysis reveals that the difference between the actual percentage of people who died vs. the values predicted by our model is centered around zero and doesn't show any pattern; this means each country is independent of one another, knowing data of one country wouldn't be enough to make a solid statement about other countries death percentage. We also found out that predictions made by our model for all countries are neither over-predicting nor under-predicting consistently, meaning that our model is good with forecasts in general. Lastly, we also found out that the residual or difference between the actual percentage of the population which died due to covid and the predicted value comes from a uniform distribution; for the most part, this is only violated when the residual or the difference is too big. That means our model prediction is good and we can know by what maximum value will our prediction vary from the actual value.

## Evaluating The Model 2

The validity of the previous analysis depends on the assumptions we have made by using a regression model, so we need to verify which assumptions we violate.

### Review of Key Takeaways

```{r}
linmod2 <- lm(covid_new$perecentage_new_deaths ~ covid_new$percentage_vaccinated + covid_new$percentage_fully_vaccinated)
b0 <- linmod2$coef[1]
b1 <- linmod2$coef[2]
b2 <- linmod2$coef[3]

Yhat <- linmod2$fitted.values
e <- resid(linmod2)
plot(Yhat,e, xlab = expression('fitted values' ~ hat(Y)),ylab="residual e", pch = 16)
abline(h = 0, lty = 3)
```

```{r}
Y1 <- covid_new$perecentage_new_deaths - b0 - b1*covid_new$percentage_vaccinated
e <- resid(linmod2)
par(mfrow=c(1,2))
plot(covid_new$percentage_fully_vaccinated,Y1, xlab = expression('% of People Fully Vaccinated'),ylab='Y_i - b_0 - b1 * % of People Vaccinated', pch = 16)
abline(h = 0, lty = 3)
Y2 <- covid_new$perecentage_new_deaths - b0 - b2*covid_new$percentage_fully_vaccinated
plot(covid_new$percentage_vaccinated, Y2, xlab = expression('% of People Vaccinated'),ylab='Y_i - b_0 - b2 * % of People Fully Vaccinated', pch = 16)
abline(h = 0, lty = 3)
```

```{r}
e <- resid(linmod2)
par(mfrow=c(2,2))
plot(covid_new$percentage_vaccinated, e, xlab="% of People Vaccinated")
plot(covid_new$percentage_fully_vaccinated, e, xlab="% of People Fully Vaccinated")
hist(e, main="Histogram of residuals", xlab="residual e", col="darkmagenta", freq=FALSE)
qqnorm(e, pch = 1, frame = FALSE)
qqline(e, col = "steelblue", lwd = 2)
```

### Independence

For this regression model, the % of People fully Vaccinated and % of People Vaccinated is independent to % of new deaths. This is because the % of vaccinated is not grouped/ aggregated by country thus there are multiple points for a single country at different day's which are making the dependence to each other very difficult. Thus the relation between the residual error is quite independent. Therefore we are not violating the independence assumption of regression models.

### Equal variance

We are also assuming equal variance in our regression model. From the residuals against the fitted values plot, we can see that they make a cone-shape, clearly, there isn't a consistent vertical spread throughout the graph. There is a huge violation of equal variance due to this inconsistent spread.

### Linearlity

There are specific regions in the graph with a majority of positive or negative residuals, therefore the model does under-predict or over-predict in a region, indicating that the data tends to be non-linear. The same can be said from the $Y_i-b_0$ against % Population of Death

In the case of $Y_i - b_0 - b_1$ \* % of People Vaccinated there is significant over predication when the % of People fully Vaccinated increases and in the case of $Y_i - b_0 - b_2$ \* % of People Fully Vaccinated there is significant under predication when the % of People fully Vaccinated increase

### Normality

The QQ-Plot allows us to visualize the normality of the errors from the data. There is a significant tail violation of normality in the QQ-Plot since the data is relatively evenly spread out across the line, when more points should be centered in the middle of the line. The tails also strafe from the line.

### Suggested transformation

transform $X_{1} -> X_{1}^{2}$ and $X_{2} -> X_{2}^{1/2}$

```{r}
linmod2.modify <- lm(covid_new$perecentage_new_deaths ~ I(covid_new$percentage_vaccinated ^ 2) + I(covid_new$percentage_fully_vaccinated ^ 0.5))
e.square <- linmod2.modify$residuals
Y.hat.square <- linmod2.modify$fitted.values
plot(Y.hat.square, e.square, pch = 16, xlab = "Fitted Values", ylab = "Residuals",
main = "Modified Model", cex=0.5, cex.lab=0.5, cex.axis=0.5, cex.main=0.5)
abline(h = 0, lty = 3)
```

```{r}
b0.modify <- linmod2.modify$coef[1]
b1.modify <- linmod2.modify$coef[2]
b2.modify <- linmod2.modify$coef[3]
Y <- covid_new$perecentage_new_deaths
par(mfrow = c(1, 2))
plot(covid_new$percentage_vaccinated^2, Y - b0.modify - b2.modify*(covid_new$percentage_fully_vaccinated^(1/2)), pch = 16, xlab = "Squared % of People Vaccinated",
ylab = expression(Y-b[0]-b[2]*X[2]^(1/2)), main = "Modified Model",
cex=0.5, cex.lab=0.5, cex.axis=0.5, cex.main=0.5)
abline(a = 0, b = b1.modify, col = "blue")
plot(covid_new$percentage_fully_vaccinated^(1/2), Y - b0.modify - b1.modify*covid_new$percentage_vaccinated^2, pch = 16, xlab = "Root % of People Fully Vaccinated",
ylab = expression(Y-b[0]-b[1]*X[1]^2), main = "Modified Model",
cex=0.5, cex.lab=0.5, cex.axis=0.5, cex.main=0.5)
abline(a = 0, b = b2.modify, col = "blue")
```

## Evaluating The Model 3

The validity of the previous analysis depends on the assumptions we have made by using a regression model, so we need to verify which assumptions we violate.

```{r}
agg_new_deaths <- aggregate(covid_new$perecentage_new_deaths, by=list(covid_new$location), mean)$x
agg_fully_vaccinated_population <- aggregate(covid_new$percentage_fully_vaccinated, by=list(covid_new$location), mean)$x
agg_vaccinated_populations <- aggregate(covid_new$percentage_vaccinated, by=list(covid_new$location), mean)$x

linmod.2 <- lm(agg_new_deaths ~ agg_vaccinated_populations + agg_fully_vaccinated_population)
b0 <- linmod.2$coef[1]
b1 <- linmod.2$coef[2]
b2 <- linmod.2$coef[3]

Yhat <- linmod.2$fitted.values
e <- resid(linmod.2)
plot(Yhat,e, xlab = expression('fitted values' ~ hat(Y)),ylab="residual e", pch = 16)
abline(h = 0, lty = 3)
```

```{r}
Y1 <- agg_new_deaths - b0 - b1*agg_vaccinated_populations
par(mfrow=c(1,2))
e <- resid(linmod.2)
plot(agg_new_deaths,Y1, xlab = expression('Aggrigated % of People Fully Vaccinated Population'),ylab='Y_i - b_0 - b1 * Aggrigated % of People Vaccinated Populations', pch = 16)
abline(h = 0, lty = 3)
Y2 <- agg_new_deaths - b0 - b2*agg_fully_vaccinated_population
plot(agg_new_deaths, Y2, xlab = expression('Aggrigated % of People Vaccinated Populations'),ylab='Y_i - b_0 - b2 * Aggrigated % of People Fully Vaccinated Population', pch = 16)
abline(h = 0, lty = 3)
```

```{r}
e <- resid(linmod.2)
par(mfrow=c(2,2))
plot(agg_fully_vaccinated_population, e, xlab="Aggrigated % of People Fully Vaccinated Population")
plot(agg_vaccinated_populations, e, xlab="Aggrigated % of People Vaccinated Population")
hist(e, main="Histogram of residuals", xlab="residual e", col="darkmagenta", freq=FALSE)
qqnorm(e, pch = 1, frame = FALSE)
qqline(e, col = "steelblue", lwd = 2)
```

### Independence

For this regression model, the % of People fully Vaccinated and % of People Vaccinated might be dependent to % of new deaths. This is because the % of vaccinated is grouped/ aggregated by country thus there is a point for a single country. And via the graph, the % of People fully Vaccinated and % of People Vaccinated have dependence to % of new deaths thus affecting the relation between the residual errors (Showing a dependence). Therefore we might be violating the independence assumption of regression models.

### Equal variance

We are also assuming equal variance in our regression model. From the residuals against the fitted values plot, we can see that they make a thin diamond shape, clearly, there is a consistent vertical spread throughout the graph. There is no violation of equal variance due to this consistent spread.

### Linearlity

There are no specific regions in the graph with a majority of positive or negative residuals, therefore the model doesn't exclusively under-predict or over-predict in a region, indicating that the data tends to be linear. The same can be said from the $Y_i - b_0 - b_1$ \* % of People Vaccinated against % Population of Death and $Y_i - b_0 - b_2$ \* % of People Fully Vaccinated against % Population of Death

### Normality

The QQ-Plot allows us to visualize the normality of the errors from the data. There is a no tail violation of normality in the QQ-Plot since the data is relatively evenly spread out across the line. The tails don't strafe much from the line. And the distribution is not "compacted" on the ends. The quantiles from residuals distribution almost match quantiles from a normal distribution. Thus assumption of normality is not violated.

### Suggested transformation

Here we don't require a lot or any transformation cause most, if not all, assumptions are not violated.
